---
title: "Pipeline Visium Processing"
output: html_document
date: "2025-04-17"
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

This script should provide a overview of the sample processing in
python. I have written addition bash scripts for automated and fast
samples processing. Here we start with Visium data from a adata format.
The processing can be done from SPATA2 as well squidpy or scanpy as a
input point. I start with the count matrix and spatial information in
the adata.uns. Here it is important to notice that if you aum to run
multiple samples in the same experiment, you should run the
normalisation in scVI with the defined samples batches. If you run only
a single sample you dont need the batch normalisation (but potentially
at later stage...).

The pipeline will contain the following steps:

1.  Adata input and preprocessing

2.  scVI normalisation and Cell2Location

3.  CytoSpce deconvolution

4.  Graph annotation

5.  Pytorch implementation (PYG outputs)

6.  GNN Training (model selection etc.)

7.  Prediction and interpretation

8.  PYG 2 single-cell graph plots

## Requirments:

I use a conda environment calles EcoFoundation which has the requirments
in the EcoFoundation.yaml file and the environment C2L c2l.yaml as well
as the environment cytospace cytospace.yaml. Unfortunatly, the pipeline
is not able to run in a single conda envioronment. In R software SPATA2
2.04 was used. All functions and required scripts are stored in the
EcoFoundation github:

# Load data and Create SPATA2 and adata object:

## Load packages

```{r}
reticulate::use_condaenv("EcoFoundation")
#reticulate::use_condaenv("c2l")
## Load Library
library(DeepSPATA)
library(SPATA2)
library(Seurat)
library(reticulate)
library(tidyverse)
library(igraph)
library(kableExtra)
library(igraph)
library(readxl)
library(anndata)

```

## Load source files

```{r}
source("/Users/henrikheiland/Desktop/RecurrentGBM/SingleCellRef/import.R")

## Import colors
pycolors <- importPyColors()

```

## Set up file system

```{r}
# Set a root:
root = "/Users/henrikheiland/Desktop/Cooperations/AQP4/Spatial"
file_system <- readRDS(paste0(root, "/file_system.RDS"))

## Define the outs folder loacations:
outs_folder = paste0(dir(root, full.names = T), "/outs")
## Define sample names
samples=paste0("sample_", dir(root, full.names = F))

## Define data.frame for file system:

file_system=data.frame(samples=samples,
                       outs = outs_folder,
                       SPATA = paste0(root,"/", samples, ".RDS"),
                       adata = paste0(root,"/", samples, ".h5ad"))


#saveRDS(file_system, paste0(root, "/file_system.RDS"))

```

## Initiate objects

```{r}

## Loop the files and return objects:
for(i in 1:nrow(file_system)){
  
  spata = SPATA2::initiateSpataObject_10X(file_system$outs[i], sample_name = file_system$samples[i])
  adata = SPATA2ANDATA(spata)
  
  ## Store
  saveRDS(spata, file_system$SPATA[i])
  anndata::write_h5ad(adata, filename = file_system$adata[i])
  
}

```

## Initialize the C2L runs

For the spatial deconv. we need the optimal type of cells. Here we use
the single cell data files from the same conditions:

```{r}
## Set up the input reference single cell data
seurat = readRDS("/Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/integrated_new.RDS")
DimPlot(seurat, group.by = "celtype_level1", label=T)
adata_sc = anndata::read_h5ad("/Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/integrated_data_new.h5ad")

## We define "batch" as the batch variable. Make sure that this exitsts in you data:
adata_sc$obs["batch"]
adata_sc$obs["celltype"] = as.character(seurat$celtype_level1)
anndata::write_h5ad(adata_sc, filename = "/Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/integrated_data_new.h5ad")

## Here are my bash: 
# cd /Users/henrikheiland/Desktop/MERFISH/EcoFoundation/R_Functions/bash_Script
#./run_cell2location_sc.sh /Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/Cell2Location /Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/integrated_data_new.h5ad
```

## Run Cell 2 location deconvolution:

In the next step we can run cell to location on all files

```{python}
import subprocess
adatas = r.file_system.adata.values
sample = r.file_system.samples.values
#Usage: Cell2location.sh <AD_FILE> <INF_CSV> <OUTPUT_CSV> [MAXEPOCHS]

# loop through all adata files
for i, adata_file in enumerate(adatas):
    infer_csv = "/Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/Cell2Location/reference_signatures/infer.csv"
    output_csv = f'{r.root}/{sample[i]}_decov.csv'
    max_epochs = "100"

    cmd = [
        "bash",
        "/Users/henrikheiland/Desktop/MERFISH/EcoFoundation/R_Functions/bash_Script/Cell2location.sh",
        adata_file,
        infer_csv,
        output_csv,
        max_epochs ]

    print("Running:", " ".join(cmd))
    subprocess.run(cmd)


```

## Plot Cell2Location plots

If you want to plot the cell2location plots you need to change the
connected conda to cell2loc:

```{r}
reticulate::use_condaenv("c2l_old")
reticulate::use_python("/Users/henrikheiland/miniconda/envs/c2l/bin/python")
root = "/Users/henrikheiland/Desktop/Cooperations/AQP4/Spatial"
file_system <- readRDS(paste0(root, "/file_system.RDS"))
```

Jump to python:

```{python}
import scanpy as sc
import numpy as np
import matplotlib.pyplot as plt
import matplotlib as mpl
import cell2location
from cell2location.plt import plot_spatial

## load files:
adatas = r.file_system.adata.values

## add .uns
import shutil
import os
for i in range(4):
  source_file = f'{(r.file_system.outs.values)[i]}/spatial/tissue_positions.csv' 
  destination_file = f'{(r.file_system.outs.values)[i]}/spatial/tissue_positions_list.csv' 
  shutil.copyfile(source_file, destination_file)
  create_ann = sc.read_visium((r.file_system.outs.values)[i])
  adata = sc.read(adatas[i])
  adata.uns['spatial'] = create_ann.uns['spatial']
  adata.obsm['spatial'] = np.asarray(create_ann.obsm['spatial'], dtype='float32')
  adata.write_h5ad(adatas[i])
  os.remove(destination_file)


adata = sc.read(adatas[3])
adata.obs[adata.uns['mod']['factor_names']] = adata.obsm['q05_cell_abundance_w_sf']
adata.obs[adata.uns['mod']['factor_names']].columns
clust_labels = ['MES_like', 'Astrocytes','MG_pro_infl','Pericyte']
clust_col = ['' + str(i) for i in clust_labels]


with mpl.rc_context({'figure.figsize': (15, 15)}):
    fig = plot_spatial(
        adata=adata, 
        color=clust_col,
        show_img=True,
        labels=clust_labels, 
        style='fast', 
        max_color_quantile=0.8,
        circle_diameter=4, 
        colorbar_position='right'
    )
plt.show()

```

Update cell2location outputs to spata objects
```{r}
## Add cell2location to the spata files:
for(i in 1:nrow(file_system)){
  object = readRDS(file_system$SPATA[i])
  sample_ID <- file_system$samples[i]
  
  message(sample_ID)
  
  cell_types_path <- paste0(root, "/",sample_ID,"_decov.csv")
  cell_types <- read.csv(cell_types_path)
  names(cell_types)[1] <- "barcodes"
  cell_types_df <- cell_types
  
  names(cell_types_df) <- str_remove(names(cell_types_df), "q05cell_abundance_w_sf_")
  names(cell_types_df) <- names(cell_types_df) %>% str_replace_all(., "[.]", "_")
  object <- object %>% SPATA2::addFeatures(cell_types_df, overwrite = T)
  saveRDS(object, file_system$SPATA[i])
}


cell_types_all <- names(cell_types_df)[2:length(names(cell_types_df))]
saveRDS(cell_types_all, paste0(root, "cell_types_all.RDS"))

```

## Run Cytospace

First you need to create the required outputs from the reference
dataset:

```{r}
source("~/Desktop/MERFISH/EcoFoundation/R_Functions/Deconvolution.R")
## First generate outputs from the single cells to run cytosypace:
## Here it is important that the celltypes are stored again in "celltype" (as seen above)
library(Seurat)
library(data.table)

scrna_seurat = readRDS("/Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/integrated_new.RDS")
scrna_seurat$celltype <- scrna_seurat$celtype_level1
folder_ref = paste0(root,"/CytoSpace/Ref")
if(!dir.exists(folder_ref)){dir.create(paste0(root,"/CytoSpace"));dir.create(folder_ref)}
generate_cytospace_from_scRNA_seurat_object(scrna_seurat, dir_out = folder_ref,  rna_assay = "RNA")

```

Second: you can run the samples in a loop: Here it is important to know
that the Deconvolution.R file creates a bash script output:

`run <- paste0("conda run -n cytospace_v1.1.0 cytospace --scRNA-path '" , Ref_mat, "' --cell-type-path '", Ref_lables, "' --st-path '", paste0(sample_dir, "/counts.txt"), "' --coordinates-path '", paste0(sample_dir, "/coords.txt"), "' -o '", sample_dir, "' --cell-type-fraction-estimation-path '", paste0(sample_dir, "/Cell_Fraction.txt"), "' --solver-method 'lap_CSPR'" )`

This will call cytospace in the defined conda env: conda run -n
cytospace_v1.1.0 cytospace. It is important to define and update to you
specific localisation and conda environment names.

Also here we use the lap_CSPR solver which based on the fact that the lapjv solver is only working in intle architectures!

```{r}
## Set parameter and run loop:
## Here it is important that the celltypes are stored again in "celltype" (as seen above)
cell_types_all <- readRDS(paste0(root, "cell_types_all.RDS"))

for(i in 1:nrow(file_system)){
  print(i)
  object = readRDS(file_system$SPATA[i])
  sample_ID <- file_system$samples[i]
  message(paste0(" 🚀 Run sample: ",sample_ID))
  Reference_Annotation <- cell_types_all
  CS_folder <- paste0(root,"/CytoSpace/", sample_ID)
  Ref_mat = paste0(root,"/CytoSpace/Ref/scRNA_data.txt")
  Ref_lables = paste0(root,"/CytoSpace/Ref/cell_type_labels.txt")
  returnCytoSpace(object, Reference_Annotation, CS_folder, Ref_mat, Ref_lables)
  
} 

shell = paste0(root,"/Cytospace/", file_system$samples, "/cytoSpace.sh")

```

```{python}
## Now the create bash script can be executed

cytospace_files = r.shell

for i in range(4):
  cmd = ["bash", cytospace_files[i]]
  print("Running:", " ".join(cmd))
  subprocess.run(cmd)

```


Import Cytospace into spata (as a single cell data.frame with spatial
coords and the annotated cell position)

```{r}

for(i in 1:nrow(file_system)){
  print(i)
  object = readRDS(file_system$SPATA[i])
  sample_ID <- file_system$samples[i]
  message(paste0("Run sample: ",sample_ID))
  
  Reference_Annotation <- cell_types_all
  CS_folder <- paste0(root,"/CytoSpace/", sample_ID)
  Ref_mat = paste0(root,"/CytoSpace/Ref/scRNA_data.txt")
  Ref_lables = paste0(root,"/CytoSpace/Ref/cell_type_labels.txt")
  
  cytospace = importCytoSpace(object, Reference_Annotation, CS_folder, Ref_mat, Ref_lables, scale=10)
  object@spatial[[1]]$cytospace <- cytospace
  saveRDS(object, file_system$SPATA[i])
  
} 

```

After you run all deconvolutionsteps you can create some nice infered single cell plots:

```{r}
## Read color schema
colors <- readRDS("/Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/colors.RDS")
object <- readRDS(file_system$SPATA[3])
celltype <- object@spatial[[1]]$cytospace
cc <- colors$colors;names(cc) <- colors$celtype_level2

library(ggforce)
plotSurface(object,color_by="Pdgfb", alpha_by="Pdgfb")

ggplot() +
    geom_voronoi_tile(data = celltype, 
                      aes(x_sc, y_sc, group = -1L, fill = CellType), max.radius = 4, colour = 'black', linewidth=0.1, alpha=0.8)+
    scale_fill_manual(values=cc)+
    theme_bw() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_rect(colour = "black", size=0.5),
          axis.text.x = element_text(colour="black"),
          axis.text.y = element_text(colour="black"))+
    coord_fixed()+
  theme(legend.text = element_text(size=6),
        legend.key.size = unit(0.3, "cm"),
        legend.title = element_text(size=8))+
  SPATA2::ggpLayerAxesSI(object)+
  xlab("Dimension x-space [cm]")+
  ylab("Dimension y-space [cm]")

```

## Graph annotation

In the next step we will process the samples and generate the graphs for
the deep learning approch.

```{python}
import sys
sys.path.append('/Users/henrikheiland/Desktop/MERFISH/EcoFoundation/PythonPackage/EcoFoundation/functions')  
from utils import *
from plt import *

import scanpy as sc
import tangram as tg
import anndata as ad
import squidpy as sq
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns

samples_all = r.file_system.samples.values

for sample in samples_all:
  print(sample)
  adata = sc.read(f"{r.root}/{sample}.h5ad")
  ## Run preprocessing
  sc.pp.pca(adata)
  sc.pp.neighbors(adata)
  sc.tl.umap(adata)
  ## scVI
  import scvi
  scvi.model.SCVI.setup_anndata(adata)
  model = scvi.model.SCVI(adata, n_latent=50)
  model.train(accelerator = "cpu", max_epochs=200, early_stopping=True)
  adata.obsm["X_scVI"] = model.get_latent_representation(adata)
  adata.layers["X_exp"] = model.get_normalized_expression(adata)
  adata.write_h5ad(f"{r.root}/{sample}.h5ad")

```


# Deep Learning approch:

## Add predictiom parameters:
We have to define what we aim to predict. Here I just use the 0/1 source of a sample and add this to our file_system:

```{r}
file_system$status = c(0,1,1,0)
```


## First we need to train a VAE which has all batches to provide normalized expression values:
In this step we remove batch effects by train the dataset together and create a common VAE model. This will be used further to normalize the gene expression when single sample graphs will be created. This is very important to remove any batch effects fro the input data!

```{python}
## Merge all files and add batch:
files = r.file_system["adata"].values
status = r.file_system["status"].values
samples_all = r.file_system["samples"].values

adata_list = []
for j,sample in enumerate(samples_all):
    print(sample)
    adata = ad.read_h5ad(f"{r.root}/{sample}.h5ad")
    adata.obs["status"] = status[j]
    adata_list.append(adata)
adata_list

for i, adata in enumerate(adata_list):
    sample_name = samples_all[i]  # Create a unique sample identifier
    adata.obs_names = [f"{bc}_{sample_name}" for bc in adata.obs_names]
    adata.obs["batch_ID"] = sample_name
adata_combined = ad.concat(adata_list, axis=0, join='outer', label='batch', keys=range(len(adata_list)), index_unique='-')
adata_combined.obs["batch"]

## process and filter
adata_combined.layers['counts'] = adata_combined.X.copy()
sc.pp.filter_cells(adata_combined, min_counts=40)
adata_combined.obs["batch"]


## Set up the model: 
scvi.model.SCVI.setup_anndata(adata_combined, batch_key='batch')

# Train:
# Train SCVI model for integration
model = scvi.model.SCVI(adata_combined, n_latent=50)
model.train(accelerator = "cpu", max_epochs=100, early_stopping=True)


# Use the trained model to get the corrected latent space
adata_combined.obsm["X_scVI"] = model.get_latent_representation()

## Save the model and adata file:
model.save(f"{r.root}/VAE_ep250_fullData")
adata_combined.write_h5ad(f"{r.root}/VAE_ep200_fullData.h5ad")
```



## Build ecosystems 
This is the key snip of code to create you ecosystems out of the combined pretrained scVI dataset. Here we build the ecosystem sample by sample. The .x part of the PYG graph object contains the information which is taken into the GNN. Here we add the full expression dataset but it is also possible to use the latentspace (n=50) or something else into the data.x output of the PYG.

```{python}

adata_combined = sc.read_h5ad(f"{r.root}/VAE_ep200_fullData.h5ad")
reference_model=scvi.model.SCVI.load(f"{r.root}/VAE_ep250_fullData",adata_combined)


## Create data into a subgraphs
## Function for loop:
## It is important that this function need to be optimized based on the requirments of the subgraphs. The Parameter and thresholds need to be optimized based on the sample requirments. Note! Differences between Visium and MERSCOPE!

#__________________________________________________________________________________________#
def runfastgraph(adata, status, batch_id):
  
  adata.layers["counts"] = adata.X
  adata.obs["batch"]=batch[batch_id]
  adata.X = reference_model.get_normalized_expression(adata)
  
  adata = BuildGraph(adata,distance_threshold = 10) #This is the threshold used to filter the distance matrix!

  adata = barcode_mapping(adata)
  # hop (number of NN), max_overlap: The maximum numbers of overlaps per subgraph, min_nodes The minimal number of nodes in a graph!
  subgraphs = getMaxSubgraphs(adata, hop=3, max_overlap=3, min_nodes=8) # Set up hyperparameters for the subgraphs
  sub = BuildPYG(adata, subgraphs)
  
  # After the baseline subgraphs are created, we add edge atributtes, status etc. This can be adjusted to whatever you plan to add!
  
  ## Get the raw data
  for i in range(len(sub)):
    node_indices = np.asarray(sub[i].node_index)
    sub[i].raw = torch.tensor(adata.layers["counts"][node_indices].toarray(), dtype=torch.float)
    sub[i].edge_attr = sub[i].distance.unsqueeze(1)
    sub[i].status = torch.tensor(status, dtype=torch.long)
    sub[i].patient_source = torch.tensor(batch[batch_id], dtype=torch.long)
  
  return sub
#__________________________________________________________________________________________#



## Here we define the specific input that are required to build the graph:
files = r.file_system["adata"].values ## File names
status = r.file_system["status"].values ## The status that will be predicted
samples_all = r.file_system["samples"].values ## Samples ID

# If the status is not encoded as integer:
t = pd.Series(status)  
t_cat = pd.Categorical(t)
status_num = t_cat.codes
# The batches 
batch = [0,1,2,3]


## Here we build the subgraph of the first batch 
graph = runfastgraph(adata_combined[adata_combined.obs['batch'] == 0].copy(),status_num[0], batch[0]).copy()

## Add graphs to the existing one:
for i in range(1,len(files)):
  gg = runfastgraph(adata_combined[adata_combined.obs['batch'] == i].copy(), status_num[i], batch[i]).copy()
  graph = graph+gg


## Check output
len(graph)
torch.save(graph, f"{r.root}/subgraphs.pt")
sub = graph

```

```{r}
saveRDS(file_system, paste0(root, "/file_system.RDS"))
```

We preprocessed the samples and build our graphs (Ecosystems). From this point we will proceed with the prediction tasks.


# Used GNN to predict:

## Load Subgraphs:
```{python}
import torch
sub = torch.load(f"{r.root}/subgraphs.pt")
```

Here we would be interested in the NN which are predictive for sample status (0 high vs 1). This will be represented by a GNN including nodes (spots) and edges (distance). We will use a attention GNN including the attentions as outputs for XAI.

Load the architectures: 

```{python}
import sys
sys.path.append('/Users/henrikheiland/Desktop/MERFISH/EcoFoundation/PythonPackage/EcoFoundation/models')  
from GAM_V1 import *
```

In the 2nd step we build the full model by adding a MLP to the GNN architecture:
```{python}

class GraphMERFISH_MultiTask(torch.nn.Module):
    def __init__(self, num_features_exp, hidden_channels, num_classes_task1, edge_dim=1):
        super().__init__()
        self.encoder = GraphEncoder(num_features_exp, hidden_channels, edge_dim=edge_dim)
        self.task1_head = MLP(hidden_channels, num_classes_task1)

    def forward(self, data):
        x, edge_index, edge_attr, batch = data.x, data.edge_index, data.edge_attr, data.batch
        node_latent, graph_latent, att1, att2 = self.encoder(x, edge_index, edge_attr, batch)
        task1_out = self.task1_head(graph_latent)
        return node_latent, task1_out, att1, att2

```

## Training function including discriminator:
```{python}
from tqdm import tqdm
import numpy as np

def train_discriminator(model, discriminator, train_loader, val_loader, criterion1, criterion_discriminator, optimizer, optimizer_discriminator, num_epochs=50, patience=5):
    early_stopping = EarlyStopping(patience=patience, delta=0.01)

    for epoch in range(num_epochs):
        model.train()
        discriminator.train()
        epoch_loss = 0
        total_discriminator_loss = 0

        for data in tqdm(train_loader):
            # Train primary model
            optimizer.zero_grad()
            data.edge_attr = data.distance.view(-1, 1).float()
            node_latent, task1_out, att1, att2 = model(data.to(device))
            
            
            #loss status
            status = data.status.long().to(device)
            loss = criterion1(task1_out, status)
            
            
            loss.backward()
            optimizer.step()
            epoch_loss += loss.item()

            # Train discriminator
            optimizer_discriminator.zero_grad()
            patient_source = data.patient_source.long().to(device)
            discriminator_pred = discriminator(node_latent.detach())
            discriminator_pred = global_mean_pool(discriminator_pred, data.batch)
            discriminator_loss = criterion_discriminator(discriminator_pred, patient_source)
            discriminator_loss.backward()
            optimizer_discriminator.step()
            total_discriminator_loss += discriminator_loss.item()

            # Adversarial step
            optimizer.zero_grad()
            node_latent, task1_out, att1, att2 = model(data.to(device))
            discriminator_pred = discriminator(node_latent)
            discriminator_pred = global_mean_pool(discriminator_pred, data.batch)
            adversarial_loss = -criterion_discriminator(discriminator_pred, patient_source)
            adversarial_loss.backward()
            optimizer.step()

        # Validation (primary model only)
        model.eval()
        val_loss = 0
        val_outputs_1 = []
        val_labels = []

        with torch.no_grad():
            for data in val_loader:
                data.edge_attr = data.distance.view(-1, 1).float()
                node_latent, task1_out, att1, att2 = model(data.to(device))
                val_outputs_1.append(torch.argmax(task1_out, dim=1).detach().cpu().numpy())
                pred_1 = data.status.long().to(device)
                val_labels.append(pred_1.cpu().numpy())
                val_loss += criterion1(task1_out, pred_1).item()

        val_loss /= len(val_loader)
        pred = np.concatenate(val_outputs_1)
        true_labels = np.concatenate(val_labels)

        # Print results
        print("Epoch {}/{}".format(epoch + 1, num_epochs))
        print("Train Loss: {:.4f}, Discriminator Loss: {:.4f}, Val Loss: {:.4f}".format(epoch_loss / len(train_loader), total_discriminator_loss / len(train_loader), val_loss))
        print("Validation Accuracy:", accuracy_score(true_labels, pred))

        # Early stopping
        early_stopping(val_loss, model)
        if early_stopping.early_stop:
            print("Early stopping")
            break

    model.load_state_dict(torch.load('checkpoint.pt'))


def train(model, train_loader, val_loader, criterion1, optimizer, num_epochs=50, patience=5):
    early_stopping = EarlyStopping(patience=patience, delta=0.01)

    for epoch in range(num_epochs):
        model.train()
        epoch_loss = 0

        for data in tqdm(train_loader):
            # Train primary model
            optimizer.zero_grad()
            data.edge_attr = data.distance.view(-1, 1).float()
            node_latent, task1_out, att1, att2 = model(data.to(device))
            
            
            #loss status
            status = data.status.long().to(device)
            loss = criterion1(task1_out, status)
            
            
            loss.backward()
            optimizer.step()
            epoch_loss += loss.item()

        # Validation (primary model only)
        model.eval()
        val_loss = 0
        val_outputs_1 = []
        val_labels = []

        with torch.no_grad():
            for data in val_loader:
                data.edge_attr = data.distance.view(-1, 1).float()
                node_latent, task1_out, att1, att2 = model(data.to(device))
                val_outputs_1.append(torch.argmax(task1_out, dim=1).detach().cpu().numpy())
                pred_1 = data.status.long().to(device)
                val_labels.append(pred_1.cpu().numpy())
                val_loss += criterion1(task1_out, pred_1).item()

        val_loss /= len(val_loader)
        pred = np.concatenate(val_outputs_1)
        true_labels = np.concatenate(val_labels)

        # Print results
        print("Epoch {}/{}".format(epoch + 1, num_epochs))
        print("Train Loss: {:.4f}, Val Loss: {:.4f}".format(epoch_loss / len(train_loader), val_loss))
        print("Validation Accuracy:", accuracy_score(true_labels, pred))

        # Early stopping
        early_stopping(val_loss, model)
        if early_stopping.early_stop:
            print("Early stopping")
            break

    model.load_state_dict(torch.load('checkpoint.pt'))

```

## Run Training:
Here we define hyperparameters and build the full model:
```{python}
device="cuda" if torch.cuda.is_available() else "cpu"
#----------------- Parameter:
num_features_exp = sub[1].x.shape[1] #Number of genes
hidden_channels=125 # Hidden channels GNN
num_classes_task1=2 # Number of classes to be predicted
edge_dim=1 #Number (dimensions) of edge information

# Discriminator:
num_classes=4 #Number of batches

model = GraphMERFISH_MultiTask(
    num_features_exp=num_features_exp, 
    hidden_channels=hidden_channels, 
    num_classes_task1=num_classes_task1, 
    edge_dim=1
)

discriminator = Discriminator(input_size=125, hidden_size=32, num_classes=num_classes) ## ! input_size ==hidden_channels

# Optimizer:
optimizer = torch.optim.Adam(model.parameters(), lr=1e-3, weight_decay=1e-4)
optimizer_discriminator = torch.optim.Adam(discriminator.parameters(), lr=0.001)

# Loss functions:
criterion1 = nn.CrossEntropyLoss()
criterion_discriminator = nn.NLLLoss()

```


## Create Dataloader and train
Here my experiance: The sparser the input data the more a model benefit from using a discriminator for batch effect removal. As soon as the model get richer inputs do not use the discriminator. Also in case your model contains already normalized and batch cleaned data, I would use only normal training:
```{python}

from torch.utils.data import random_split
from sklearn.metrics import accuracy_score, precision_score, recall_score, f1_score, roc_auc_score, confusion_matrix

## Prepare training and val data:
data_list = sub
train_ratio = 0.8
val_ratio = 0.2
train_size = int(len(data_list) * train_ratio)
val_size = len(data_list) - train_size
train_data, val_data = random_split(data_list, [train_size, val_size])

# Create data loaders
from torch_geometric.loader import DataLoader
train_loader = DataLoader(train_data, batch_size=10, shuffle=False)
val_loader = DataLoader(val_data, batch_size=10, shuffle=False)

# If you ant to test:
#data = next(iter(train_loader))
#data.edge_attr = data.distance.view(-1, 1).float()
#first_batch = next(iter(train_loader))
#node_latent, task1_out, att1, att2 = model(data.to(device))

## With discriminator:
train_discriminator(
    model=model,
    discriminator=discriminator,
    train_loader=train_loader,
    val_loader=val_loader,
    criterion1=criterion1,
    criterion_discriminator=criterion_discriminator,
    optimizer=optimizer,
    optimizer_discriminator=optimizer_discriminator,
    num_epochs=50,
    patience=5
)

## Without discriminator:
train(
    model=model,
    train_loader=train_loader,
    val_loader=val_loader,
    criterion1=criterion1,
    optimizer=optimizer,
    num_epochs=50,
    patience=5
)

## Store model:
torch.save(model, f"{r.root}/GNN_trained_AQP4_KO_noDis.pt")

```

Finally we have trained our model. Now its time to look mor in detail and evaluate the predicted outputs. Here I used 4 samples which are samples from either WT or APQ4-KO mouse with a PDGFB tumor. We tried to determine Ecosystems for WT or AQP4 tumors:

#Evaluate data:

```{python}

model.eval()
val_outputs_1 = []
val_labels1 = []
logit1 = []


full_data = DataLoader(sub, batch_size=10, shuffle=False)

with torch.no_grad():
  for data in full_data:
    data.edge_attr = data.distance.view(-1, 1).float()
    node_latent, task1_out, att1, att2 = model(data.to(device))
    logit1.append(task1_out.detach().cpu().numpy())
    val_outputs_1.append(torch.argmax(task1_out, dim=1).detach().cpu().numpy())
    pred_1 = data.status.long().to(device)
    val_labels1.append(pred_1.cpu().numpy())

pred_1_all = np.concatenate(val_outputs_1)
label_1_all = np.concatenate(val_labels1)
logits_1_all = np.concatenate(logit1)


## Check confusion map
predicted = pred_1_all
labels = label_1_all
print("Accuracy:", accuracy_score(labels, predicted))
print("Precision:", precision_score(labels, predicted, average='macro'))
print("Recall:", recall_score(labels, predicted, average='macro'))
print("F1 Score:", f1_score(labels, predicted, average='macro'))
print("Confusion Matrix:\n", confusion_matrix(labels, predicted))

cm = confusion_matrix(labels, predicted)
import seaborn as sns
import matplotlib.pyplot as plt
plt.figure(figsize=(10, 8))
sns.heatmap(cm, annot=True, fmt='g', cmap='Blues')
plt.xlabel('Predicted')
plt.ylabel('True')
plt.title('Confusion Matrix: True vs Predicted')
plt.show()

```
In this example a clear overfitting on the data but we aim to get some XAI features:

```{r}
# Get all the results in a dataframe:
res = data.frame(status = py$pred_1_all, logit = py$logits_1_all)
saveRDS(res, paste0(root, "/", "prediction_res.RDS"))


## Plot the logits:
## If you have multiple classes you can also use the logits as an input for a umap! (Nice plots)

library(scales)
ggplot(res)+
  geom_point(mapping = aes(x=logit.1 %>% rescale(., c(0,1)), 
                           y=logit.2 %>% rescale(., c(0,1)),
                           color = logit.1 %>% rescale(., c(0,1))),
             )+
  scale_color_gradientn(colors = rev(RColorBrewer::brewer.pal(9, "RdBu")))+
  theme_bw() +
      theme(panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.background = element_rect(colour = "black", size=0.5),
            axis.text.x = element_text(colour="black"),
            axis.text.y = element_text(colour="black"))+
      coord_fixed()+
    theme(legend.text = element_text(size=6),
          legend.key.size = unit(0.3, "cm"),
          legend.title = element_text(size=8))+
    guides(color = guide_colourbar(barwidth = 0.3, barheight = 8, ticks =F, frame.colour="black"), label=F)


```

## Analysis of representative ecosystem
Here it is important to know that we do not uses the same downstream analysis between Visium and MERFISH data (based on the resolution). So this example uses Visium and our deconvolution data to demonstrate the dwonstream analysis. I also don't want to go into much detail. Here is the way to get typical enriched ecosystems on single cell level:

```{r}
## Read SPATA object
spata_obj <-  readRDS(file_system$SPATA[1])
## Get Cell type deconvolution
celltype = spata_obj@spatial[[1]]$cytospace
```

Next we get all subgraphs, define the inputs and function:
```{python}
#__________________________________________________________________________________________#
def runfullgraph(adata, status, batch_id):
  
  adata.layers["counts"] = adata.X
  adata.obs["batch"]=batch[batch_id]
  adata.X = reference_model.get_normalized_expression(adata)
  
  adata = BuildGraph(adata,distance_threshold = 5) #This is the threshold used to filter the distance matrix!
  adata = barcode_mapping(adata)
  # Here we allow overlap!
  subgraphs, central_nodes = getfullSubgraphs(adata, hop=3, min_nodes=20) # Set up hyperparameters for the subgraphs
  sub = BuildPYG(adata, subgraphs)
  # After the baseline subgraphs are created, we add edge atributtes, status etc. This can be adjusted to whatever you plan to add!
  
  ## Get the raw data
  for i in range(len(sub)):
    node_indices = np.asarray(sub[i].node_index)
    sub[i].raw = torch.tensor(adata.layers["counts"][node_indices].toarray(), dtype=torch.float)
    sub[i].edge_attr = sub[i].distance.unsqueeze(1)
    sub[i].status = torch.tensor(status, dtype=torch.long)
    sub[i].patient_source = torch.tensor(batch[batch_id], dtype=torch.long)
  
  return sub, central_nodes
#__________________________________________________________________________________________#


#__________________________________________________________________________________________#
## Here we define the specific input that are required to build the graph:
files = r.file_system["adata"].values ## File names
status = r.file_system["status"].values ## The status that will be predicted
samples_all = r.file_system["samples"].values ## Samples ID
# If the status is not encoded as integer:
t = pd.Series(status)  
t_cat = pd.Categorical(t)
status_num = t_cat.codes
# The batches 
batch = [0,1,2,3]
#__________________________________________________________________________________________#

```

Run evaluation:
```{python}
adata_combined = sc.read_h5ad(f"{r.root}/VAE_ep200_fullData.h5ad")
adata_eval = adata_combined[adata_combined.obs['batch'] == 0].copy()
graph, nodes = runfullgraph(adata_eval,status_num[0], batch[0])

model.eval()
val_outputs_1 = []
val_labels1 = []
logit1 = []
bcs_subgraph = []

with torch.no_grad():
  for i, data in enumerate(graph):
    data.edge_attr = data.distance.view(-1, 1).float()
    node_latent, task1_out, att1, att2 = model(data.to(device))
    logit1.append(task1_out.detach().cpu().numpy())
    val_outputs_1.append(torch.argmax(task1_out, dim=1).detach().cpu().numpy())
    pred_1 = data.status.long().to(device)
    val_labels1.append(pred_1.cpu().numpy())
    bcs_subgraph.append(adata_eval.obs.index[data.node_index].values)

label_1_all = np.hstack(val_labels1)
logits_1_all = np.concatenate(logit1)
bcs_all = adata_eval.obs.index[nodes].values

```

Merge the information together in a data.frame
```{r}
res = 
  data.frame(subgraph_nr = 1:length(py$bcs_all), barcodes_graph = py$bcs_all, status = py$pred_1_all, logit = py$logits_1_all) %>% 
  mutate(barcodes = barcodes_graph %>% str_split(., "_") %>% map(.,~.x[[1]]) %>% unlist()) %>% 
  mutate(KO = logit.1)
subgraph_nodes <- map(py$bcs_subgraph,~ .x %>% str_split(., "_") %>% map(.,~.x[[1]]) %>% unlist())

```

Here we can get the full prediction and plot it over the images:
```{r}
spata_obj <- addFeatures(spata_obj, data.frame(barcodes=res$barcodes, KO = res$KO), overwrite = T)
SPATA2::plotSurface(spata_obj, pt_alpha=0.5, color_by = "KO", limits=c(4.9,5), oob=scales::squish)
```


```{r}
## Get top ecosystems for KO prediction
res %>% arrange(desc(KO)) %>% head(20)

subgraph_nr=2136
## Read color schema
colors <- readRDS("/Users/henrikheiland/Desktop/Cooperations/AQP4/SingleCell/colors.RDS")
cc <- colors$colors;names(cc) <- colors$celtype_level2
plotEcosystem(spata_obj, subgraph_nodes, subgraph_index=subgraph_nr, extend=0.2)

```






















